Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph\tnoteref{t1,t2}

### 2017-01-20 response_to_reviewers_2.txt

Anonymous comments from Referee #2:

Page 12, line 23 [page 12, line 25]:
I would still suggest adding the word bits to make it clearer, because the Big-O notation is also used with the unit 'words'.
* We revised the sentence to say "... using O(n) space (in bits).".

Page 12, line 52-55 [page 12, line 57]:
The correction is fine, but in the new last sentence on page 12, 'four' should probably be 'for'.
* Corrected.

Page 16, lines 45-46 [page 16, lines 43-46]:
The added explanation is helpful. However, I am not sure whether it is obvious enough that the "access times distributions do not change substantially" through the exclusions. Maybe this also requires some proof and a more precise statement of the effect (compared to just saying "substantial").
* We apologize for the misunderstanding. This figure is simply meant to visually demonstrate the effect of curvature on access times when all other variables are held constant. The excluded pairs have much smaller access times than those shown but which are not substantially different (visually) from those of the trees from which they can be obtained via replacement. We were not claiming that the excluded trees have the same access times as the included trees, and, as such there is nothing to prove here. As the replaced trees have a smaller number of leaves (and therefore much smaller access times) the effect is a messy plot with multiple curves that essentially includes the results of trees with 3 to say 7 leaves (for (b)). Each individual curve shows a similar influence from curvature.
* We revised these sentences to say: "To limit clutter, these plots exclude the pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings). The additional curves induced by the excluded pairs show a similar influence from curvature."

Page 20, Lemma 5.5, point 1 [page 20, Lemma 5.5]:
Unfortunately, the revised sentence introducing the cases is not much more helpful than its predecessor (for me). The main source of confusion is this: "we do not need to exclude the target branch from k and a, gaining an additional leaf to balance the product ka at the cost of increasing j" (from the case where the subtree is moved to the root). Probably I am missing a point somewhere, but I cannot make much sense of this sentence. Exclude the target branch from k and a? Gaining an additional leaf - where / how?
* We apologize for the confusion, which was likely caused by our incorrect definition of a and b in Lemma 5.3. In the special case that R is moved to an ancestor, the newly introduced node becomes L in the new tree S. Thus, the node "L" of T should also be part of the path that defines a, because moves of the subtree rooted at "L" can now be attached to R. In the case described in Lemma 5.5 we thus increase the length of a by 1 when we move R to the root. We originally defined (a - b) in terms of the height change experienced by R but felt the path-based definition was more intuitive and thereby made this error. Thus, this does not change any of our results.
* We have revised the statement and proof of Lemma 5.3, introducing separate least common ancestors L_1 in T and L_2 in S, defining a using the path from U to L_2, and defining b using the path from V to L_1.
* We have revised the proof of Lemma 5.5 to state:

"Recall that k is the number of leaves in the moved subtree R, i is the number of leaves in R’s neighbor U and j is the number of leaves in R’s new neighbor V after the SPR move. The length of the path from U ’s root to the least common ancestor L 2 of U and V in S is a, while b is the length of the path from V ’s root to the least common ancestor L 1 of U and V in T . Call the former path A.

The degree change is maximized by making L_1 the root of T and thereby minimizing b, namely setting b = 0. The resulting equation 2(ka + i − j) is similarly maximized by including only one leaf in V, setting i = 1. We then maximally balance the terms in the product ka by adding as many leaves as possible to either R or as children of the path A. Each new leaf of R increases k by 1 and each new child of A lengthens the path and increases a by 1.

There are two cases, depending on whether R is moved to the root (thereby introducing a new level to the tree) or not. If not, then the j leaves of the new neighbor subtree V cannot contribute to increasing k or a. We thus set j = 1 and split the remaining n − b − i − j = n − 2 leaves between R and A in as balanced a way as possible, maximizing the product ka and giving (i). Note that this corresponds to moving the bottom subtree of floor((n−2)/2) or ceiling((n-2)/2) leaves in a ladder tree to the root-most leaf of the tree.

If R is moved to the root, then the new neighbor subtree V is the entire tree excluding R itself. The root-most leaf of the tree is on the path from U to L in S and thus a child of A, despite also being a child of L in T . In this case, we have an additional leaf over the previous case that can be added to R or lengthen A, increasing the product ka at the cost of increasing j. This corresponds to moving the bottom subtree of floor(n/2)) or ceiling(n/2) leaves in a ladder tree to the root. ..."

### 2017-01-20 curvature.tex
addressed suggestion

	\hspace*{\stretch{1}}
	\subfigure[\label{fig:5t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-long-time-kappa-access}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:6t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
	\hspace*{\stretch{1}}
    \caption{\
        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-leaf trees with degree 44 and (b) 7-leaf trees with degree 70.
        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
				To limit clutter, these plots exclude the pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings). These additional curves show a similar influence from curvature.
        Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
				See online version for color figures; data tables available at \protect\url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
        }
	\label{fig:kappa-access}
\end{figure*}

@@

\begin{lem}
\label{lem:degree_change}
Let $T$ and $S$ be trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

\end{lem}
\begin{proof}
The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
(i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
(ii) subtrees that include nodes on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
(iii) $R$'s parent subtree may now be moved into $U$, and
(iv) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked by the original rSPR operation on $R$.

@@

\begin{lem}
	\label{lem:degree_max_delta_adjacent}
	Let $T$, $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{lem}
\begin{proof}
	We first prove (i).
	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) and then maximally balancing the terms in the product $ka$ as follows.

	There are two cases, depending on whether the subtree of $k$ leaves is moved to the root (thereby introducing a new level to the tree) or not.
    If not, then the $j$ leaves of the new neighbor subtree $V$ cannot be part of $k$ or $a$. We thus set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $k$ and $a$ in as balanced a way as possible, giving (i).
	Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the root-most leaf of the tree.

	If the subtree of $k$ leaves is moved to the root, then the new neighbor subtree $V$ is the entire tree excluding $R$ itself. We can thus include the root-most leaf of the tree in $a$. In this case, we gain an additional leaf over the previous case to balance the product $ka$ at the cost of increasing $j$.
	This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
	Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
	Let ${\Delta}N = \degree{S} - \degree{T}$.
	If we move the additional leaf, we have:
	\begin{align*}
		{\Delta}N &\le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) \\
		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
	\end{align*}
like before.
Similarly, if we do not move the additional leaf, we also have:
\begin{align*}
	{\Delta}N &\le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) \\
&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
\end{align*}
proving (i).

>>>>

	\hspace*{\stretch{1}}
	\subfigure[\label{fig:5t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/5t-long-time-kappa-access}}
	\hspace*{\stretch{2}}
	\subfigure[\label{fig:6t-long-time-kappa-access}]{\includegraphics[width=2.3in]{figs/6t-long-time-kappa-access}}
	\hspace*{\stretch{1}}
    \caption{\
        Distribution of early rSPR $\MH$ access times for those pairs of (a) 6-leaf trees with degree 44 and (b) 7-leaf trees with degree 70.
        Distribution of later rSPR $\MH$ access times for those pairs of (c) 5-leaf trees with degree 24 and (d) 6-leaf trees with degree 44.
				To limit clutter, these plots exclude the pairs that result from modifying pairs of trees with a smaller number of leaves by replacing a given leaf with a given subtree (e.g\ we exclude pairs of trees that both contain leaves 1 and 2 as siblings). The additional curves induced by the excluded pairs show a similar influence from curvature.
        Color signifies rSPR distance between the trees, with green, orange, blue, and light blue signifying distances of 1, 2, 3, and 4, respectively; the saturation of the color shows coarse curvature $\curvature{\MH; \cdot, \cdot}$, such that increased saturation (i.e.\ darker color) indicates a smaller $\kappa$.
				See online version for color figures; data tables available at \protect\url{https://github.com/matsengrp/curvature/blob/master/prefigs/figure-5-6ab-data.zip}.
        }
	\label{fig:kappa-access}
\end{figure*}

@@

\begin{lem}
\label{lem:degree_change}
Let $T$ and $S$ be trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L_1$ be the $\lca{U,V}$ in $T$.
	Let $L_2$ be the $\lca{U,V}$ in $S$ (typically $L_1 = L_2$).
Let $a$ be the number of intermediate nodes on the path from $U$ to $L_2$ in $S$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L_1$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

\end{lem}
\begin{proof}
The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
(i) subtrees that include nodes on the path from $U$ to $L_2$ may now be moved into $R$ and its newly introduced parent node,
(ii) subtrees that include nodes on the path from $V$ to $L_1$ may no longer be moved into $R$ and its parent node,
(iii) $R$'s parent subtree may now be moved into $U$, and
(iv) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked by the original rSPR operation on $R$.

@@

\begin{lem}
	\label{lem:degree_max_delta_adjacent}
	Let $T$, $S$ be trees with $n \ge 3$ leaves, s.t. $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{lem}
\begin{proof}
We first prove (i).
By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
Recall that $k$ is the number of leaves in the moved subtree $R$, $i$ is the number of leaves in $R$'s neighbor $U$ and $j$ is the number of leaves in $R$'s new neighbor $V$ after the SPR move.
The length of the path from $U$'s root to the least common ancestor $L_2$ of $U$ and $V$ in $S$ is $a$, while $b$ is the length of the path from $V$'s root to the least common ancestor $L_1$ of $U$ and $V$ in $T$.
Call the former path $A$.

The degree change is maximized by making $L_1$ the root of T and thereby minimizing $b$, namely setting $b=0$.
The resulting equation $2(ka + i - j)$ is similarly maximized by including only one leaf in $V$, setting $i=1$. We then maximally balance the terms in the product $ka$ by adding as many leaves as possible to either $R$ or as children of the path $A$. Each new leaf of $R$ increases $k$ by 1 and each new child of $A$ lengthens the path and increases $a$ by 1.

There are two cases, depending on whether $R$ is moved to the root (thereby introducing a new level to the tree) or not.
If not, then the $j$ leaves of the new neighbor subtree $V$ cannot contribute to increasing $k$ or $a$. We thus set $j=1$ and split the remaining $n-b-i-j = n-2$ leaves between $R$ and $A$ in as balanced a way as possible, maximizing the product $ka$ and giving (i).
Note that this corresponds to moving the bottom subtree of $\floor{\frac{n-2}{2}}$ or $\ceil{\frac{n-2}{2}}$ leaves in a ladder tree to the root-most leaf of the tree.

If $R$ is moved to the root, then the new neighbor subtree $V$ is the entire tree excluding $R$ itself. The root-most leaf of the tree is on the path from $U$ to $L$ in $S$ and thus a child of $A$, despite also being a child of $L$ in $T$. In this case, we have an additional leaf over the previous case that can be added to $R$ or lengthen $A$, increasing the product $ka$ at the cost of increasing $j$.
This corresponds to moving the bottom subtree of $\floor{\frac{n}{2}}$ or $\ceil{\frac{n}{2}}$ leaves in a ladder tree to the root.
Namely, we have $2(ka + 1 - j)$, where $j = n - k = a + 1$.
Let ${\Delta}N = \degree{S} - \degree{T}$.
If we move the additional leaf, we have:
	\begin{align*}
		{\Delta}N &\le 2\left(\ceilXXL{\frac{n}{2}}\floorV{\frac{n-2}{2}}  + 1 - \left(\floorV{\frac{n-2}{2}} + 1\right)\right) \\
		&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
	\end{align*}
like before.
Similarly, if we do not move the additional leaf, we also have:
\begin{align*}
	{\Delta}N &\le 2 \left(\ceilV{\frac{n-2}{2}}\floorXXL{\frac{n}{2}} +1 -  \left(\ceilV{\frac{n-2}{2}} + 1\right)\right) \\
&= 2\floorV{\frac{n-2}{2}}\ceilV{\frac{n-2}{2}},
\end{align*}
proving (i).
